Deep ordinal classification based on cumulative link models

### 2019-05-29 paper/ieee.tex
Paper, wiki

	\author{Víctor-Manuel~Vargas, Pedro-Antonio~Gutiérrez and César~Hervás}

@@

	% The paper headers
	\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
	{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for IEEE Journals}
	% The only time the second header will appear is for the odd numbered pages
	% after the title page when using the twoside option.
	% 
	% *** Note that you probably will NOT want to include the author's ***
	% *** name in the headers of peer review papers.                   ***
	% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
	% you desire.

@@

	% As a general rule, do not put math, special symbols or citations
	% in the abstract or keywords.
	\begin{abstract}
		This paper proposes a deep convolutional neural network model for ordinal regression by considering a family of probabilistic ordinal link functions in the output layer. The link functions are those used for cumulative link models, which are traditional statistical linear models based on projecting each pattern into a 1-dimensional space. A set of ordered thresholds splits this space into the different classes of the problem. In our case, the projections are estimated by a non-linear deep neural network. To further improve the results, we combine these ordinal models with a loss function that takes into account the distance between the categories, based on the weighted Kappa index. Three different link functions are studied in the experimental study, and the results are contrasted with statistical analysis. The experiments run over two different ordinal classification problems and the statistical tests confirm that these models improve the results of a nominal model and outperform other proposals considered in the literature.
	\end{abstract}

	% Note that keywords are not normally used for peerreview papers.
	\begin{IEEEkeywords}
		Deep learning, ordinal regression, cumulative link models.
	\end{IEEEkeywords}

@@

	\section{Introduction}
	\label{sect:introduction}
	% The very first letter is a 2 line initial drop letter followed
	% by the rest of the first word in caps.
	% 
	% form to use if the first word consists of a single letter:
	% \IEEEPARstart{A}{demo} file is ....
	% 
	% form to use if you need the single drop letter followed by
	% normal text (unknown if ever used by the IEEE):
	% \IEEEPARstart{A}{}demo file is ....
	% 
	% Some journals put the first two words in caps:
	% \IEEEPARstart{T}{his demo} file is ....
	% 
	% Here we have the typical use of a "T" for an initial drop letter
	% and "HIS" in caps to complete the first word.
	\IEEEPARstart{D}{eep learning}, introduced by Yann Lecun~\cite{lecun2015deep}, combines multiple machine learning techniques and allows computational models that are composed of numerous processing layers to learn representations of data with various levels of abstraction. These methods have dramatically improved the state-of-the-art in many domains, such as image classification~\cite{cirecsan2012multi,he2016deep,krizhevsky2012imagenet}, speech recognition~\cite{hinton2012deep}, control problems~\cite{mnih2015human}, object detection~\cite{jiang2016speed,girshick2014rich}, privacy and security protection~\cite{yu2017iprivacy,yuan2019adversarial,tan2018deep}, recovery of human pose~\cite{hong2015multimodal}, semantic segmentation~\cite{long2015fully} and image retrieval~\cite{li2015weakly,jin2018deep}. Convolutional Neural Networks (CNN) are one of the types of deep networks that are designed to process data that come in the form of multiple arrays. CNNs are appropriate for images, video, speech and audio processing, and they have been used extensively in the last years for automatic classification tasks~\cite{dong2014learning,sun2013deep,ronneberger2015u}. For images, each colour channel is represented by a 2D array, and convolutional layers extract the main features from the pixels, and, after that, a fully connected layer classify every sample based on its extracted features. At the output of the CNN, a softmax function provides the probabilities of the set of classes predefined in the model for classification tasks. However, the softmax may not be the best option depending on the problem considered.

@@

	In this paper, we propose the use of CLMs for deriving deep learning ordinal classifiers. In the case of CNNs, the model projection used by the threshold model can be obtained from the last layer of the network. Given that we work with a 1-dimensional space, the last layer would have only one neuron (projection of the pattern), and its value could be used to classify the sample into the corresponding class according to the thresholds. Some previous works have used the \texttt{logit} in shallow neural networks~\cite{gutierrez2016ordinal}, but this strategy has not been considered for deep learning, and alternative link functions have not been evaluated. To further improve the results, we train these models by minimising an ordinal loss function based on the Weighted Kappa index \cite{de2018weighted}, instead of using the standard cross-entropy.

@@

	%\textbf{(Métrica ordinal (MAE))}
	M. ALALI et al. \cite{alali2018multi} proposed a complex CNN architecture for solving Twitter Sentiment Classification as an ordinal problem. They checked that using average pooling preserves significant features that provide more expressiveness to ordinal scale. They didn't propose any method to include the ordinal information into the classifier, but they tried to find the best CNN model architecture based on an ordinal metric.

@@

	Z. Niu et al. \cite{niu2016ordinal} proposed a learning approach to address ordinal regression problems using CNNs. They divided the problem into a series of binary classification sub-problems and proposed a multiple output CNN optimization algorithm to collectively solve these classification sub-problems, taking into account the correlation between them.

	H. Li et al. \cite{li2017deep} applied deep learning techniques for solving the ordinal problem of Alzheimer's diagnosis and detecting the different levels of the disease as multiple binary sub-problems.

	Y. Liu et al. \cite{liu2017deep} proposed a new approach which transforms the ordinal regression problem to binary classification sub-problems and use triplets with instances from different categories to train deep neural networks. In this way, high-level features describing the ordinal relationships are extracted automatically. Given that triplets must be generated, this approach is only recommended for small datasets.

	S. Chen et al. \cite{chen2017using} proposed a deep learning method termed Ranking-CNN. This method combines multiple binary CNNs that are trained with ordinal age labels. The binary outputs are aggregated for the final age prediction. They achieved a tighter error bound for ranking-based age estimation.

@@

	J. de la Torre et al.~\cite{de2018weighted} proposed the use of a continuous version of the quadratic weighted kappa (QWK) metric as loss function for the optimization algorithm. They compared this cost function against the traditional log-loss function using three different datasets, including the Diabetic Retinopathy database as the most complex one. They proved that their function could improve the results as it reduces overfitting and training time. Also, they checked the importance of hyper-parameter tuning.

	A. Rios et al. \cite{rios2017ordinal} presented a CNN model designed to handle ordinal regression tasks on psychiatric notes. They combined an ordinal loss function, a CNN model and conventional feature engineering. Also, they applied a technique called Locally Interpretable Model-agnostic Explanation (LIME) to make the non-linear model more interpretable.

	H. Fu et al. \cite{fu2018deep} applied deep learning techniques to Monocular Depth Estimation. They introduced a spacing-increasing discretization strategy to treat the problem as an ordinal regression problem. They improved the performance when training the network with an ordinal regression loss. Also, they used a multi-scale network structure that avoids unnecessary spatial pooling.

	A. Pal et al. \cite{pal2018severity} defined a loss function for CNN that is based on the Earth Mover's Distance and takes into account the ordinal class relationships.

	Y. Liu et al. \cite{liu2018constrained} proposed a constrained optimization formulation for the ordinal regression problem which minimizes the negative loglikelihood for a multi-class problem constrained by the order relationship between instances.

@@

	C. Beckham and C. Pal~\cite{beckham2017unimodal} proposed a straightforward technique to constrain discrete ordinal probability distributions to be unimodal, via the use of the Poisson and binomial probability distributions. The parameters of these distributions were learnt by using a deep neural network. They evaluated this approach on two large ordinal image datasets, including the Adience dataset used in this paper, obtaining promising results. They also included a simple squared-error reformulation \cite{beckham2016simple} that was sensitive to class ordering.

@@

	\subsubsection{Adience}
	Adience\footnote{http://www.openu.ac.il/home/hassner/Adience/data.html} dataset consists of $26580$ faces belonging to $2284$ subjects. We use the form of the dataset where faces have been pre-cropped and aligned. The dataset was preprocessed, using the methods described in a previous work~\cite{beckham2017unimodal}, so that the images are 256 pixels in width and height, and pixels values follow a $(0;1)$ normal distribution. The original dataset was split into five cross-validation folds. The training set consists of merging the first four folds which comprise a total of $15554$ images. From this, $10\%$ of the images are held out as part of a validation set. The last fold is used as test set. Some images of this dataset are shown in Figure \ref{fig:AdienceExamples}. Adience dataset has been used in other works for human age estimation but most of them solved the problem as a multi-class problem instead of using the ordinal relation between classes. E. Eidinger \cite{eidinger2014age} presented an approach using support vector machines and neural networks. J.-C. Chen \cite{chen2016cascaded} proposed a coarse-to-fine strategy for deep CNNs. G. Levi \cite{levi2015age} presented another convolutional network model for age estimation, while Beckham \cite{beckham2017unimodal} proposed a straightforward method to constrain discrete ordinal probability distributions to be unimodal.

@@

	\begin{table}[!t]
		\caption{Description of the architecture used for each dataset.}
		\label{table:Architecture}
		\centering
		\resizebox{\columnwidth}{!}{
		\begin{tabular}{lclc}
			\hline\hline
			\multicolumn{2}{c}{\textbf{Diabetic Retinopathy}} & \multicolumn{2}{c}{\textbf{Adience}}\\
			\hline
			\textbf{Layer}  & \textbf{Output} & \textbf{Layer}      & \textbf{Output} \\ \hline
			2 Conv3x3@32s1  &   252x252x32    & Conv7x7@32s2        &   112x112x32    \\
			MaxPool2x2s2    &   126x126x32    & MaxPool3x3s2        &    55x55x32     \\
			2 Conv3x3@64s1  &   122x122x64    & 2 ResBlock3x3@64s1  &    55x55x32     \\
			MaxPool2x2s2    &    61x61x64     & ResBlock3x3@128s2   &    28x28x64     \\
			2 Conv3x3@128s1 &    57x57x128    & 2 ResBlock3x3@128s1 &    28x28x64     \\
			MaxPool2x2s2    &    28x28x128    & 1 ResBlock3x3@256s2 &    14x14x128    \\
			2 Conv3x3@128s1 &    24x24x128    & 2 ResBlock3x3@256s1 &    14x14x128    \\
			MaxPool2x2s2    &   12x12x128\    & 1 ResBlock3x3@512s2 &     7x7x256     \\
			Conv4x4@128s1   &     9x9x128     & 2 ResBlock3x3@512s1 &     7x7x256     \\
			                &                 & AveragePool7x7s2    &     1x1x256     \\ \hline\hline
		\end{tabular}}	
	\end{table}

@@

		\item Accuracy-based metrics. Correct Classification Rate (CCR) or standard accuracy is the most common metric for classification tasks and shows the percentage of correctly classified samples. We also include Top-2 CCR \cite{beckham2017unimodal} and Top-3 CCR \cite{beckham2017unimodal}, which are similar to CCR, but they take a prediction as correct when the real class is between the two or three classes, respectively, with the highest probability.

@@

	\subsection{Factors}
	In our study, three different factors are considered:
	\begin{itemize}
		\item \textit{Learning rate} (LR, $\eta$). LR is one of the most critical hyper-parameters to tune for training deep neural networks. Optimal learning rate can vary depending on the dataset and the CNN architecture. Previous works have presented some techniques that adjust this parameter in order to achieve better performance~\cite{smith2017cyclical,senior2013empirical}. In this work, we consider three different values for the initial value of this parameter: $10^{-2}$, $10^{-3}$ and $10^{-4}$.
		\item \textit{Batch size} (BS). Batch size is also an important parameter as it controls the number of weight updates that are made on every epoch. It can affect the training time and the model performance. In this paper, we try three different batch sizes for each dataset. For the DR dataset, we use $5$, $10$ and $15$, while, for Adience, $64$, $128$ and $256$ images are used. We took the batch sizes that were used in \cite{de2018weighted} and \cite{beckham2017unimodal} as a reference, and we expand the range on both sides.
		\item \textit{Link function} (LF). Different link functions are used for the CLM at the last layer output: \texttt{logit}, \texttt{probit} and complementary log-log (see Section \ref{sect:ordinalproblem}).
	\end{itemize}

@@

	\begin{table}[!t]
		\caption{Tukey's test results for both datasets.}
		\label{table:Tukey}
		\centering
		\resizebox{\columnwidth}{!}{
		\begin{tabular}{cccccc}
			\hline\hline
			                  &                   & \multicolumn{2}{c}{\textbf{DR}} & \multicolumn{2}{c}{\textbf{Adience}} \\ \hline
			       LF         &        LF         & Mean diff. &        Sig.        & Mean diff. &          Sig.           \\ \hline
			 \texttt{logit}   &  \texttt{probit}  &  $0.046$   &      $0.000$       &  $-0.002$  &         $0.011$         \\
			                  & \texttt{clog-log} &  $0.084$   &      $0.000$       &  $0.012$   &         $0.000$         \\
			 \texttt{probit}  &  \texttt{logit}   &  $-0.046$  &      $0.000$       &  $0.002$   &         $0.011$         \\
			                  & \texttt{clog-log} &  $0.038$   &      $0.000$       &  $0.014$   &         $0.248$         \\
			\texttt{clog-log} &  \texttt{logit}   &  $-0.084$  &      $0.000$       &  $-0.012$  &         $0.000$         \\
			                  &  \texttt{probit}  &  $-0.038$  &      $0.000$       &  $-0.014$  &         $0.248$         \\ \hline\hline
			       LR         &        LR         & Mean diff. &        Sig.        & Mean diff. &          Sig.           \\ \hline
			    $10^{-2}$     &     $10^{-3}$     &  $-0.046$  &      $0.000$       &  $-0.073$  &         $0.000$         \\
			                  &     $10^{-4}$     &  $0.148$   &      $0.000$       &  $-0.044$  &         $0.000$         \\
			    $10^{-3}$     &     $10^{-2}$     &  $0.046$   &      $0.000$       &  $0.073$   &         $0.000$         \\
			                  &     $10^{-4}$     &  $0.194$   &      $0.000$       &  $0.029$   &         $0.023$         \\
			    $10^{-4}$     &     $10^{-2}$     &  $-0.148$  &      $0.000$       &  $0.044$   &         $0.000$         \\
			                  &     $10^{-3}$     &  $-0.194$  &      $0.000$       &  $-0.029$  &         $0.023$         \\ \hline\hline
			       BS         &        BS         & Mean diff. &        Sig.        & Mean diff. &          Sig.           \\ \hline
			       64         &        128        &  $-0.041$  &      $0.026$       &     -      &            -            \\
			                  &        256        &  $-0.027$  &      $0.000$       &     -      &            -            \\
			       128        &        64         &  $0.041$   &      $0.026$       &     -      &            -            \\
			                  &        256        &  $0.014$   &      $0.000$       &     -      &            -            \\
			       256        &        64         &  $0.027$   &      $0.000$       &     -      &            -            \\
			                  &        128        &  $-0.014$  &      $0.000$       &     -      &            -
		\end{tabular}}
	\end{table}

@@

	To sum up, the results showed that the best parameter configuration depends on the problem that is being solved. The optimal value for the batch size and the optimal LF are not the same for Retinopathy and Adience datasets. These results highlight the importance of adjusting the hyper-parameters for each problem instead of trying to find an optimal configuration for all the datasets. However, the best LR for both datasets were $10^{-3}$. It is recommended to use this value for future datasets. The best BS for DR was 10, while the best value for Adience was 128 (intermediate values considered). Finally, there are more interactions between the three factors for the Adience dataset than for DR. This highlights the importance of making experimental designs associated with each dataset to determine the best value for each factor.

@@

	These results are included in Table \ref{table:Comparison}, together with the comparison against previous works of the state-of-the-art. All the results are given for the test set, except those from \cite{de2018weighted} (DR dataset), because the authors only provided validation results for $128\times 128$ images (however, validation results are usually better than test results). The results for \cite{beckham2017unimodal} were obtained by reproducing their experiments because they didn't provide exact results for test set that could be used to compare with our results. The proposed ordinal model outperforms all the other alternatives in terms of QWK.

@@

	\begin{table}[!t]
		\caption{Comparison between the best results of nominal, ordinal and previous works for the both datasets.}
		\label{table:Comparison}
		\scriptsize
		\centering
		\def\arraystretch{1.3}
		\begin{tabular}{lccc}
			\hline\hline
			\multicolumn{4}{c}{\textbf{Diabetic Retinopathy}}\\
			\hline
			Method                                    & $\overline{QWK}_{(SD)}$ & $\overline{CCR}_{(SD)}$ & $\overline{\text{1-off}}_{(SD)}$ \\ \hline
			Ordinal network                           &    $0.582_{(0.016)}$    &    $0.723_{(0.004)}$    &        $0.868_{(0.002)}$         \\
			Nominal network                           &    $0.498_{(0.013)}$    &    $0.692_{(0.014)}$    &        $0.854_{(0.006)}$         \\
			J. Torre et al. \cite{de2018weighted}     &  $0.537_{(\text{-})}$   &            -            &                -                 \\
			À. Nebot et al. \cite{nebot2016diabetic}  &  $0.555_{(\text{-})}$   &            -            &                -                 \\
			\hline
			\multicolumn{4}{c}{\textbf{Adience}}\\
			\hline
			Method                                    & $\overline{QWK}_{(SD)}$ & $\overline{CCR}_{(SD)}$ & $\overline{\text{1-off}}_{(SD)}$ \\ \hline
			Ordinal network                           &    $0.881_{(0.005)}$    &    $0.519_{(0.013)}$    &        $0.894_{(0.005)}$         \\
			Nominal network                           &    $0.787_{(0.004)}$    &    $0.458_{(0.008)}$    &        $0.800_{(0.007)}$         \\
			Beckham et al. \cite{beckham2017unimodal} &    $0.855_{(0.012)}$    &    $0.467_{(0.019)}$    &        $0.867_{(0.011)}$         \\
			E. Eidinger et al. \cite{eidinger2014age} &            -            &    $0.451_{(0.026)}$    &        $0.807_{(0.011)}$         \\
			J.-C. Chen et al. \cite{chen2016cascaded} &            -            &    $0.529_{(0.060)}$    &        $0.885_{(0.022)}$         \\
			G. Levi et al. \cite{levi2015age}         &            -            &    $0.507_{(0.051)}$    &        $0.847_{(0.022)}$         \\ \hline\hline
		\end{tabular}
	\end{table}

>>>>

	\author{Víctor-Manuel~Vargas, Pedro-Antonio~Gutiérrez,~\IEEEmembership{Senior~Member,~IEEE,} and César~Hervás-Mart\'inez,~\IEEEmembership{Senior~Member,~IEEE}}

@@

	% The paper headers
	\markboth{}%
	{}
	% The only time the second header will appear is for the odd numbered pages
	% after the title page when using the twoside option.
	% 
	% *** Note that you probably will NOT want to include the author's ***
	% *** name in the headers of peer review papers.                   ***
	% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
	% you desire.

@@

	% As a general rule, do not put math, special symbols or citations
	% in the abstract or keywords.
	\begin{abstract}
		This paper proposes a deep convolutional neural network model for ordinal regression by considering a family of probabilistic ordinal link functions in the output layer. The link functions are those used for cumulative link models, which are traditional statistical linear models based on projecting each pattern into a 1-dimensional space. A set of ordered thresholds splits this space into the different classes of the problem. In our case, the projections are estimated by a non-linear deep neural network. To further improve the results, we combine these ordinal models with a loss function that takes into account the distance between the categories, based on the weighted Kappa index. Three different link functions are studied in the experimental study, and the results are contrasted with statistical analysis. The experiments run over two different ordinal classification problems, and the statistical tests confirm that these models improve the results of a nominal model and outperform other proposals considered in the literature.
	\end{abstract}

	% Note that keywords are not normally used for peerreview papers.
	\begin{IEEEkeywords}
		Deep learning, ordinal regression, cumulative link models, Kappa index.
	\end{IEEEkeywords}

@@

	\section{Introduction}
	\label{sect:introduction}
	% The very first letter is a 2 line initial drop letter followed
	% by the rest of the first word in caps.
	% 
	% form to use if the first word consists of a single letter:
	% \IEEEPARstart{A}{demo} file is ....
	% 
	% form to use if you need the single drop letter followed by
	% normal text (unknown if ever used by the IEEE):
	% \IEEEPARstart{A}{}demo file is ....
	% 
	% Some journals put the first two words in caps:
	% \IEEEPARstart{T}{his demo} file is ....
	% 
	% Here we have the typical use of a "T" for an initial drop letter
	% and "HIS" in caps to complete the first word.
	\IEEEPARstart{D}{eep learning}, introduced by Lecun~et al. \cite{lecun2015deep}, combines multiple machine learning techniques and allows computational models that are composed of numerous processing layers to learn representations of data with various levels of abstraction. These methods have dramatically improved the state-of-the-art in many domains, such as image classification~\cite{cirecsan2012multi,he2016deep,krizhevsky2012imagenet}, speech recognition~\cite{hinton2012deep}, control problems~\cite{mnih2015human}, object detection~\cite{jiang2016speed,girshick2014rich}, privacy and security protection~\cite{yu2017iprivacy,yuan2019adversarial,tan2018deep}, recovery of human pose~\cite{hong2015multimodal}, semantic segmentation~\cite{long2015fully} and image retrieval~\cite{li2015weakly,jin2018deep}. Convolutional Neural Networks (CNN) are one of the types of deep networks that are designed to process data that come in the form of multiple arrays. CNNs are appropriate for images, video, speech and audio processing, and they have been used extensively in the last years for automatic classification tasks~\cite{dong2014learning,sun2013deep,ronneberger2015u}. For images, each colour channel is represented by a 2D array, and convolutional layers extract the main features from the pixels, and, after that, a fully connected layer classify every sample based on its extracted features. At the output of the CNN, a softmax function provides the probabilities of the set of classes predefined in the model for classification tasks. However, the softmax may not be the best option depending on the problem considered.

@@

	In this paper, we propose the use of CLMs for deriving deep learning ordinal classifiers\footnote{The source code is available at \url{https://github.com/ayrna/deep-ordinal-clm}.}. In the case of CNNs, the model projection used by the threshold model can be obtained from the last layer of the network. Given that we work with a 1-dimensional space, the last layer would have only one neuron (projection of the pattern), and its value could be used to classify the sample into the corresponding class according to the thresholds. Some previous works have used the \texttt{logit} in shallow neural networks~\cite{gutierrez2016ordinal}, but this strategy has not been considered for deep learning, and alternative link functions have not been evaluated. To further improve the results, we train these models by minimising an ordinal loss function based on the Weighted Kappa index \cite{de2018weighted}, instead of using the standard cross-entropy.

@@

	%\textbf{(Métrica ordinal (MAE))}
	ALALI et al. \cite{alali2018multi} proposed a complex CNN architecture for solving Twitter Sentiment Classification as an ordinal problem. They checked that using average pooling preserves significant features that provide more expressiveness to ordinal scale. They didn't propose any method to include the ordinal information into the classifier, but they tried to find the best CNN model architecture based on an ordinal metric.

@@

	Niu et al. \cite{niu2016ordinal} proposed a learning approach to address ordinal regression problems using CNNs. They divided the problem into a series of binary classification sub-problems and proposed a multiple output CNN optimization algorithm to collectively solve these classification sub-problems, taking into account the correlation between them.

	Li et al. \cite{li2017deep} applied deep learning techniques for solving the ordinal problem of Alzheimer's diagnosis and detecting the different levels of the disease as multiple binary sub-problems.

	Liu et al. \cite{liu2017deep} proposed a new approach which transforms the ordinal regression problem to binary classification sub-problems and use triplets with instances from different categories to train deep neural networks. In this way, high-level features describing the ordinal relationships are extracted automatically. Given that triplets must be generated, this approach is only recommended for small datasets.

	Chen et al. \cite{chen2017using} proposed a deep learning method termed Ranking-CNN. This method combines multiple binary CNNs that are trained with ordinal age labels. The binary outputs are aggregated for the final age prediction. They achieved a tighter error bound for ranking-based age estimation.

@@

	de la Torre et al.~\cite{de2018weighted} proposed the use of a continuous version of the quadratic weighted kappa (QWK) metric as loss function for the optimization algorithm. They compared this cost function against the traditional log-loss function using three different datasets, including the Diabetic Retinopathy database as the most complex one. They proved that their function could improve the results as it reduces overfitting and training time. Also, they checked the importance of hyper-parameter tuning.

	Rios et al. \cite{rios2017ordinal} presented a CNN model designed to handle ordinal regression tasks on psychiatric notes. They combined an ordinal loss function, a CNN model and conventional feature engineering. Also, they applied a technique called Locally Interpretable Model-agnostic Explanation (LIME) to make the non-linear model more interpretable.

	Fu et al. \cite{fu2018deep} applied deep learning techniques to Monocular Depth Estimation. They introduced a spacing-increasing discretization strategy to treat the problem as an ordinal regression problem. They improved the performance when training the network with an ordinal regression loss. Also, they used a multi-scale network structure that avoids unnecessary spatial pooling.

	Pal et al. \cite{pal2018severity} defined a loss function for CNN that is based on the Earth Mover's Distance and takes into account the ordinal class relationships.

	Liu et al. \cite{liu2018constrained} proposed a constrained optimization formulation for the ordinal regression problem which minimizes the negative loglikelihood for a multi-class problem constrained by the order relationship between instances.

@@

	Beckham and Pal~\cite{beckham2017unimodal} proposed a straightforward technique to constrain discrete ordinal probability distributions to be unimodal, via the use of the Poisson and binomial probability distributions. The parameters of these distributions were learnt by using a deep neural network. They evaluated this approach on two large ordinal image datasets, including the Adience dataset used in this paper, obtaining promising results. They also included a simple squared-error reformulation \cite{beckham2016simple} that was sensitive to class ordering.

@@

	\subsubsection{Adience}
	Adience\footnote{http://www.openu.ac.il/home/hassner/Adience/data.html} dataset consists of $26580$ faces belonging to $2284$ subjects. We use the form of the dataset where faces have been pre-cropped and aligned. The dataset was preprocessed, using the methods described in a previous work~\cite{beckham2017unimodal}, so that the images are 256 pixels in width and height, and pixels values follow a $(0;1)$ normal distribution. The original dataset was split into five cross-validation folds. The training set consists of merging the first four folds which comprise a total of $15554$ images. From this, $10\%$ of the images are held out as part of a validation set. The last fold is used as test set. Some images of this dataset are shown in Figure \ref{fig:AdienceExamples}. Adience dataset has been used in other works for human age estimation but most of them solved the problem as a multi-class problem instead of using the ordinal relation between classes. Eidinger et al. \cite{eidinger2014age} presented an approach using support vector machines and neural networks. Chen et al. \cite{chen2016cascaded} proposed a coarse-to-fine strategy for deep CNNs. Levi and Hassner \cite{levi2015age} presented another convolutional network model for age estimation. As previously discussed, Beckham and Pal \cite{beckham2017unimodal} proposed a straightforward method to constrain discrete probability distributions to be unimodal.

@@

	\begin{table}[!t]
		\caption{Description of the architecture used for each dataset.}
		\label{table:Architecture}
		\centering
		\resizebox{\columnwidth}{!}{
		\begin{tabular}{lclc}
			\hline\hline
			\multicolumn{2}{c}{\textbf{Diabetic Retinopathy}} & \multicolumn{2}{c}{\textbf{Adience}}  \\ \hline
			\textbf{Layer}  &         \textbf{Output}         & \textbf{Layer}      & \textbf{Output} \\ \hline
			2 Conv3x3@32s1  &           124x124x32            & Conv7x7@32s2        &   112x112x32    \\
			MaxPool2x2s2    &            62x62x32             & MaxPool3x3s2        &    55x55x32     \\
			2 Conv3x3@64s1  &            58x58x64             & 2 ResBlock3x3@64s1  &    55x55x32     \\
			MaxPool2x2s2    &            29x29x64             & ResBlock3x3@128s2   &    28x28x64     \\
			2 Conv3x3@128s1 &            25x25x128            & 2 ResBlock3x3@128s1 &    28x28x64     \\
			MaxPool2x2s2    &            12x12x128            & 1 ResBlock3x3@256s2 &    14x14x128    \\
			2 Conv3x3@128s1 &             8x8x128             & 2 ResBlock3x3@256s1 &    14x14x128    \\
			MaxPool2x2s2    &             4x4x128             & 1 ResBlock3x3@512s2 &     7x7x256     \\
			Conv4x4@128s1   &             1x1x128             & 2 ResBlock3x3@512s1 &     7x7x256     \\
			                &                                 & AveragePool7x7s2    &     1x1x256     \\ \hline\hline
		\end{tabular}}	
	\end{table}

@@

		\item Accuracy-based metrics. Correct Classification Rate (CCR) or standard accuracy is the most common metric for classification tasks and shows the percentage of correctly classified samples. We also include Top-2 CCR and Top-3 CCR \cite{beckham2017unimodal}, which are similar to CCR, but they take a prediction as correct when the real class is between the two or three classes, respectively, with the highest probability.

@@

	In order to ease reproducibility, the source code is available in a public repository\footnote{\url{https://github.com/ayrna/deep-ordinal-clm}}.
	\subsection{Factors}
	In our study, three different factors are considered:
	\begin{itemize}
		\item \textit{Learning rate} (LR, $\eta$). LR is one of the most critical hyper-parameters to tune for training deep neural networks. Optimal learning rate can vary depending on the dataset and the CNN architecture. Previous works have presented some techniques that adjust this parameter in order to achieve better performance~\cite{smith2017cyclical,senior2013empirical}. In this work, we consider three different values for the initial value of this parameter: $10^{-2}$, $10^{-3}$ and $10^{-4}$.
		\item \textit{Batch size} (BS). Batch size is also an important parameter as it controls the number of weight updates that are made on every epoch. It can affect the training time and the model performance. In this paper, we try three different batch sizes for each dataset. For the DR dataset, we use $5$, $10$ and $15$, while, for Adience, $64$, $128$ and $256$ images are used. We took the batch sizes that were used in \cite{de2018weighted} and \cite{beckham2017unimodal} as a reference, and we expand the range on both sides.
		\item \textit{Link function} (LF). Different link functions are used for the CLM at the last layer output: \texttt{logit}, \texttt{probit} and complementary log-log (see Section \ref{sect:ordinalproblem}).
	\end{itemize}

@@

	\begin{table}[!t]
		\caption{Tukey's test results for both datasets.}
		\label{table:Tukey}
		\centering
		\resizebox{\columnwidth}{!}{
		\begin{tabular}{cccccc}
			\hline\hline
			                  &                   & \multicolumn{2}{c}{\textbf{DR}} & \multicolumn{2}{c}{\textbf{Adience}} \\ \hline
			       LF         &        LF         & Mean diff. &        Sig.        & Mean diff. &          Sig.           \\ \hline
			 \texttt{logit}   &  \texttt{probit}  &  $0.046$   &      $0.000$       &  $-0.002$  &         $0.011$         \\
			                  & \texttt{clog-log} &  $0.084$   &      $0.000$       &  $0.012$   &         $0.000$         \\
			 \texttt{probit}  &  \texttt{logit}   &  $-0.046$  &      $0.000$       &  $0.002$   &         $0.011$         \\
			                  & \texttt{clog-log} &  $0.038$   &      $0.000$       &  $0.014$   &         $0.248$         \\
			\texttt{clog-log} &  \texttt{logit}   &  $-0.084$  &      $0.000$       &  $-0.012$  &         $0.000$         \\
			                  &  \texttt{probit}  &  $-0.038$  &      $0.000$       &  $-0.014$  &         $0.248$         \\ \hline\hline
			       LR         &        LR         & Mean diff. &        Sig.        & Mean diff. &          Sig.           \\ \hline
			    $10^{-2}$     &     $10^{-3}$     &  $-0.046$  &      $0.000$       &  $-0.073$  &         $0.000$         \\
			                  &     $10^{-4}$     &  $0.148$   &      $0.000$       &  $-0.044$  &         $0.000$         \\
			    $10^{-3}$     &     $10^{-2}$     &  $0.046$   &      $0.000$       &  $0.073$   &         $0.000$         \\
			                  &     $10^{-4}$     &  $0.194$   &      $0.000$       &  $0.029$   &         $0.023$         \\
			    $10^{-4}$     &     $10^{-2}$     &  $-0.148$  &      $0.000$       &  $0.044$   &         $0.000$         \\
			                  &     $10^{-3}$     &  $-0.194$  &      $0.000$       &  $-0.029$  &         $0.023$         \\ \hline\hline
			       BS         &        BS         & Mean diff. &        Sig.        & Mean diff. &          Sig.           \\ \hline
			       64         &        128        &  $-0.041$  &      $0.026$       &     -      &            -            \\
			                  &        256        &  $-0.027$  &      $0.000$       &     -      &            -            \\
			       128        &        64         &  $0.041$   &      $0.026$       &     -      &            -            \\
			                  &        256        &  $0.014$   &      $0.000$       &     -      &            -            \\
			       256        &        64         &  $0.027$   &      $0.000$       &     -      &            -            \\
			                  &        128        &  $-0.014$  &      $0.000$       &     -      &            -\\\hline\hline
		\end{tabular}}
	\end{table}

@@

	To sum up, the results showed that the best parameter configuration depends on the problem that is being solved. The \texttt{clog-log} function offers the best results in DR dataset while the \texttt{logit} link is the best option for the Adience dataset. However, the best LR for both datasets were $10^{-3}$. It is recommended to use this value for future datasets. The best BS for DR was 10, while the best value for Adience was 128 (intermediate values considered). Finally, there are more interactions between the three factors for the Adience dataset than for DR. These results highlight the importance of adjusting the hyper-parameters for each problem instead of trying to find an optimal configuration for all the datasets.

@@

	These results are included in Table \ref{table:Comparison}, together with the comparison against previous works of the state-of-the-art. All the results are given for the test set, except those from \cite{de2018weighted} (DR dataset), because the authors only provided validation results for $128\times 128$ images (however, validation results are usually better than test results). The results for \cite{beckham2017unimodal} were obtained by reproducing their experiments because they did not provide the results for test set. As can be checked, the proposed ordinal model outperforms all the other alternatives in terms of QWK.

@@

	\begin{table}[!t]
		\caption{Comparison between the best results of the CLM network, a nominal network (using softmax and cross-entropy) and previous works for both datasets.}
		\label{table:Comparison}
		\scriptsize
		\centering
		\def\arraystretch{1.3}
		\begin{tabular}{lccc}
			\hline\hline
			\multicolumn{4}{c}{\textbf{Diabetic Retinopathy}}\\
			\hline
			Method                                    & $\overline{QWK}_{(SD)}$ & $\overline{CCR}_{(SD)}$ & $\overline{\text{1-off}}_{(SD)}$ \\ \hline
			CLM network (proposal)                          &    $0.582_{(0.016)}$    &    $0.723_{(0.004)}$    &        $0.868_{(0.002)}$         \\
			Nominal network                           &    $0.498_{(0.013)}$    &    $0.692_{(0.014)}$    &        $0.854_{(0.006)}$         \\
			de la Torre et al. \cite{de2018weighted}     &  $0.537_{(\text{-})}$   &            -            &                -                 \\
			Nebot et al. \cite{nebot2016diabetic}  &  $0.555_{(\text{-})}$   &            -            &                -                 \\
			\hline
			\multicolumn{4}{c}{\textbf{Adience}}\\
			\hline
			Method                                    & $\overline{QWK}_{(SD)}$ & $\overline{CCR}_{(SD)}$ & $\overline{\text{1-off}}_{(SD)}$ \\ \hline
			CLM network (proposal)                          &    $0.881_{(0.005)}$    &    $0.519_{(0.013)}$    &        $0.894_{(0.005)}$         \\
			Nominal network                           &    $0.787_{(0.004)}$    &    $0.458_{(0.008)}$    &        $0.800_{(0.007)}$         \\
			Beckham and Pal \cite{beckham2017unimodal} &    $0.855_{(0.012)}$    &    $0.467_{(0.019)}$    &        $0.867_{(0.011)}$         \\
			Eidinger et al. \cite{eidinger2014age} &            -            &    $0.451_{(0.026)}$    &        $0.807_{(0.011)}$         \\
			Chen et al. \cite{chen2016cascaded} &            -            &    $0.529_{(0.060)}$    &        $0.885_{(0.022)}$         \\
			Levi and Hassner \cite{levi2015age}         &            -            &    $0.507_{(0.051)}$    &        $0.847_{(0.022)}$         \\ \hline\hline
		\end{tabular}
	\end{table}
